---
title: "Tampa Bay Coastal Master Plan Habitat Assessments: Habitat Evolution Model Workflow Description"
author: "Tampa Bay Estuary Program"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    theme: cosmo
    code-tools: 
      source: https://github.com/tbep-tech/tbcmp-analyze/blob/main/docs/hem-workflow.qmd
    css: styles.css
    embed-resources: true
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
# Load packages as needed
# library(tidyverse)
# library(knitr)
# library(kableExtra)
```

---

**Prepared by:** Tampa Bay Estuary Program  
263 13th Ave. S., Suite 350  
St. Petersburg, FL 33701

**Submitted to TBRPC:** February 2026

**Repository:** <https://github.com/tbep-tech/tbcmp-analyze>

---

## 1. Purpose and Scope

This document describes the analytical workflow implemented in the tbcmp-analyze GitHub repository, which serves as the primary codebase supporting the Tampa Bay Coastal Master Plan (TBCMP) Habitat Assessments. The workflow is designed to model and quantify coastal habitat vulnerability and change across the seven-county TBCMP project area---spanning Citrus, Hernando, Hillsborough, Manatee, Pasco, Pinellas, and Sarasota counties---under multiple sea level rise (SLR) scenarios and projection horizons (2050, 2080, and 2100).

The repository houses the converted and refined version of the Habitat Evolution Model (HEM), which extends the Sea Level Affecting Marsh Model (SLAMM v. 6.2) framework. The original model was implemented in ArcGIS ModelBuilder and ArcPython (ArcMap Toolbox v10.7; [Sheehan et al. 2015](https://drive.google.com/file/d/14xhQpdvlXMjUz-BG_zVS_eg2QYYZVdWz/view?usp=drivesdk)). The workflow described herein represents a complete translation of that model to the open-source R programming environment, enabling reproducible, version-controlled, and automated analyses consistent with TBEP's open science data management practices ([Beck et al. 2021](https://drive.google.com/file/d/1vO4B8DJATgCSV1qOxZz-kN6Uj1BrgNsg/view?usp=sharing)).

## 2. Repository Architecture and Data Organization

The tbcmp-analyze repository is organized following standard R project conventions, with clearly delineated directories for raw data, processed data, analytical scripts, and documentation outputs. The repository is publicly accessible on the TBEP GitHub group page and maintained under version control using Git, with all commits, issues, and changes tracked transparently.

### 2.1 Directory Structure

The repository contains four primary directories that collectively support the full analytical lifecycle from raw data acquisition through final reporting:

- **R/** --- Contains all analytical scripts constituting the Habitat Evolution Model workflow.
- **data-raw/** --- Stores unprocessed source data files downloaded or otherwise acquired from external agencies, including raster inputs and tabular parameter files.
- **data/** --- Contains processed intermediate and final data products generated by the workflow scripts.
- **docs/** --- Hosts rendered HTML reporting outputs, including the publicly accessible [HEM Land Use Change Summary](https://tbep-tech.github.io/tbcmp-analyze/hemchange.html), which provides county-level tabular summaries of projected habitat transitions given two land management scenarios, two accretion scenarios and three SLR scenarios.

Version control procedures are applied using Git software, with the repository maintaining the complete record of workflow changes, dataset updates, and analytical refinements over the project lifecycle. Identified issues and required corrections are tracked via GitHub's issue tracking system, documenting responsible parties, affected files, and implementation timelines, consistent with the protocols described in the [TBCMP Technical Strategy](https://docs.google.com/document/d/1ol1cWYR-7QZDhsDrg3hRf3xOvu_7JMQs/edit?usp=sharing&ouid=107198491532889175628&rtpof=true&sd=true).

## 3. Primary Data Inputs

The workflow relies exclusively on publicly available, pre-validated datasets from established federal and state agencies, consistent with the data acquisition protocols outlined in the [TBCMP Technical Strategy](https://docs.google.com/document/d/1ol1cWYR-7QZDhsDrg3hRf3xOvu_7JMQs/edit?usp=sharing&ouid=107198491532889175628&rtpof=true&sd=true). All datasets are retrieved programmatically using scripted routines that extract data from web endpoints, agency application programming interfaces (APIs), or downloaded through a web browser when programmatic access is unavailable. No new primary data collection is performed under this project.

### 3.1 Required Input Datasets

The following datasets are required to execute the Habitat Evolution Model:

- **Sea Level Rise Scenario Table (.csv):** Contains Year and SLR columns specifying projected SLR values in meters for each modeled year. SLR scenarios are derived from the NOAA Sea Level Rise Calculator for [St. Petersburg, FL (Station ID: 8726520)](https://coast.noaa.gov/sealevelcalculator/).
- **Tidal Datums Table (.csv):** Contains tidal benchmark elevations (HAT, MHHW, MHW, MTL, MLHW, MLW, MLLW) sourced from NOAA tidal station records for [St. Petersburg, FL (Station ID: 8726520)](https://tidesandcurrents.noaa.gov/stationhome.html?id=8726520).
- **Vegetation/Habitat Raster:** An integer-coded raster representing current habitat and land use classifications, processed at 2-meter resolution across the seven-county project area. The primary source is the 2023 Southwest Florida Water Management District (SWFWMD) land use land cover dataset and seagrass (and other subtidal habitat) coverages.
- **Digital Elevation Model (DEM):** An elevation raster in meters (NAVD88) at 2-meter resolution, co-registered with the habitat raster. NoData values are assigned a value of -999. The [FGDL FLDEM statewide DEM](https://fgdl.org/ords/r/prod/fgdl-current/catalog) coverage was merged with [NOAA's continuously updated DEM](https://www.ncei.noaa.gov/products/coastal-elevation-models) to derive a seamless DEM coverage across the subtidal and uplands of the seven county project area. 
- **Freshwater Polygon (optional but recommended):** A polygon layer derived from the [Florida National Hydrography Dataset (NHD) major river](https://geodata.dep.state.fl.us/datasets/6d08dcb218364608988457fc1aeab395_5/explore?location=28.139900%2C-84.864200%2C5) features, used to differentiate freshwater-influenced areas from saltwater-dominated zones.

### 3.2 Habitat Classification System

The workflow employs a 23-category habitat classification scheme based on codes established in the [TBEP Habitat Master Plan (Robinson et al. 2020)](https://drive.google.com/file/d/1Hp0l_qtbxp1JxKJoGatdyuANSzQrpL0I/view?usp=drivesdk). Habitats are organized into five broad groups: Developed (1000-series), Agricultural (2000-series), Upland Natural (1900, 3000--4000-series), Open Water (5000-series), Wetland (6000-series), and Coastal/Submerged (7000 and 9000-series). Key habitat types relevant to the vulnerability assessment include Mangroves (6120), Salt Marsh / High Marsh (6420), Juncus Marsh (6425), Tidal Flat (6510), Salt Barren (6600), Beach (7100), and Seagrass (9113).

## 4. Workflow Components and Script Descriptions

The Habitat Evolution Model is implemented across seven primary R scripts, each responsible for a discrete analytical function. The scripts are modular by design, allowing individual components to be executed independently for testing and validation, or sequentially through the master orchestration script for full model runs([02_HabitatEvolutionModel_v202512.R](https://github.com/tbep-tech/tbcmp-analyze/blob/main/R/02_HabitatEvolutionModel_v202512.R)). Table 1 provides a summary of each script and its outputs.

**Table 1.** *Summary of Habitat Evolution Model scripts, functions, and primary outputs.*

```{r}
#| label: tbl-scripts
#| tbl-cap: "Summary of Habitat Evolution Model scripts, functions, and primary outputs."

scripts_df <- data.frame(
  Script = c(
    "01_data_prep.R",
    "DatumAdjustment_v202512.R (helper Script)",
    "MarshAccretion_v202512.R (helper script)",
    "Ocean2Beach_v202512.R (helper script)",
    "ProcessFreshwater_v202512.R (helper script)",
    "HabitatAdjustment_v202512.R (Parts 1–3, helper scripts)",
    "HabitatEvolutionModel_v202512.R (Main script)"
  ),
  Function = c(
    "Assembles and reprojects base GIS datasets; truncates files to 7-county bounding box",
    "Adjusts tidal datums (HAT, MHHW, MTL, etc.) for each SLR scenario",
    "Calculates vertical marsh accretion using constant or elevation-dependent (Marsh98) methods",
    "Creates 500 m buffer around ocean/water bodies to identify beach conversion zones",
    "Generates freshwater influence mask using NHD waterbody polygons",
    "Core habitat reclassification logic; applies elevation thresholds, tidal datums, and land protection policies across 23 habitat types",
    "Master script; iterates sub-models through each SLR scenario and projection year"
  ),
  `Key Outputs` = c(
    "Local .tif raster files",
    "Adjusted datum values per year",
    "marshac1_{year}.tif",
    "oceancheck.tif",
    "freshwater2.tif",
    "habitat_{year}.tif per SLR scenario",
    "Final habitat evolution rasters by SLR year"
  ),
  check.names = FALSE
)

knitr::kable(scripts_df, align = c("l", "l", "l"))
```

### 4.1 01_data_prep.R --- Data Preparation

The data preparation script assembles the full suite of base GIS datasets required by the Habitat Evolution Model. Using the R packages `curl` and `tidycensus`, datasets are acquired programmatically from known agency web endpoints where feasible. All raster datasets are spatially constrained to a bounding box encompassing the seven-county coastal project area, inclusive of jurisdictional coastal waters, and saved as local GeoTIFF (.tif) files for subsequent model use. Metadata documentation for each dataset is maintained following the Ecological Metadata Language (EML) standard, consistent with TBEP data management procedures ([Beck et al. 2021](https://drive.google.com/file/d/1vO4B8DJATgCSV1qOxZz-kN6Uj1BrgNsg/view?usp=sharing)).

### 4.2 DatumAdjustment_v202512.R --- Tidal Datum Adjustment Helper Script

This script reads the tidal datums and SLR scenario tables and calculates adjusted tidal benchmark elevations for each modeled projection year. The seven tidal datums (HAT, MHHW, MHW, MTL, MLHW, MLW, MLLW) are each incremented by the corresponding SLR value, effectively translating the vertical reference frame of the model forward through time. These adjusted datums are passed to the habitat reclassification module to ensure that habitat-elevation relationships are evaluated against the appropriate tidal benchmarks for each future scenario.

### 4.3 MarshAccretion_v202512.R --- Marsh Vertical Accretion Helper Script

The marsh accretion module calculates the annual vertical accumulation of sediment and organic matter for tidal wetland habitats, adjusting the DEM surface accordingly before habitat transitions are evaluated. The script supports two accretion methodologies: a constant-rate approach and an elevation-dependent approach, which scales accretion rates based on inundation frequency. Habitat-specific accretion rates are configurable parameters, with default low values of 3.75 mm/year for Juncus marsh, 1.6 mm/year for salt marsh, and 1.6 mm/year for mangroves. High accretion scenarios use 4.0 mm/year for Juncus marsh, 3.0 mm/year for high salt marsh, and 5.0 mm/year for mangroves.  The accreted topography surface is saved as an intermediate raster (`marshac1_{year}.tif`) and used as input to the habitat adjustment module.

### 4.4 Ocean2Beach_v202512.R --- Ocean Proximity Analysis Helper Script

This script identifies areas proximate to open ocean or large water bodies that may be subject to beach formation or dune migration under sea level rise. A 500-meter buffer is generated around ocean and estuarine water bodies using the `sf` package's `st_buffer()` function. The resulting ocean proximity mask (`oceancheck.tif`) is applied during habitat reclassification to enable appropriate conversion of low-elevation coastal areas to beach and dune habitats where conditions are met.

### 4.5 ProcessFreshwater_v202512.R --- Freshwater Influence Mask Helper Script

The freshwater processing script creates a binary raster mask (`freshwater2.tif`) delineating areas under freshwater influence, based on the Florida NHD waterbody polygon layer. Pixels within NHD-defined waterbody features receive a value of 1 (freshwater), while those outside receive a value of 0 (saltwater/estuarine). This differentiation is critical to the habitat reclassification logic, as freshwater influence determines eligibility for specific habitat types: Juncus marsh and tidal flat formation are enabled in freshwater zones, while mangrove establishment is restricted to saltwater-influenced areas.

### 4.6 HabitatAdjustment_v202512.R --- Core Habitat Reclassification Helper Script

The habitat adjustment module constitutes the core scientific logic of the Habitat Evolution Model. Implemented across three script parts due to the complexity of the decision tree, this module reclassifies each habitat pixel based on its adjusted elevation relative to the SLR-modified tidal datums, its proximity to water bodies, its freshwater influence status, and applicable land development protection policies.

The elevation-based habitat zonation follows a hierarchical tidal datum framework:

- Areas above the Highest Astronomical Tide (HAT) + SLR threshold are classified as Upland or Salt Barren, depending on proximity to tidal influence.
- Areas between MHHW and HAT are classified as Salt Barren.
- Areas between MHW and MHHW are classified as High Marsh (salt marsh, 6420).
- Areas between MTL and MHW are classified as Low Marsh (Juncus marsh, 6425, in freshwater zones; High Marsh in saltwater zones).
- Areas between MLHW and MTL are classified as Tidal Flat (6510) in freshwater zones; mangrove (6120) in saltwater zones.
- Areas between MLLW and MLHW support mangrove or Tidal Flat habitats, depending on freshwater influence.
- Areas below MLLW are classified as subtidal Open Water (5400) or Seagrass (9113) where conditions support submerged aquatic vegetation.

Land development protection policies are configurable via the `Protect_Developed` parameter. When protection is enabled, all developed land use codes (1100, 1200, 1800, 1820, 2100, 2200, 2400, 2550) are preserved in place regardless of inundation status, reflecting the assumption that coastal engineering and infrastructure protection will prevent habitat migration into any developed areas. When protection is disabled, only "softly" developed lands above the HAT + SLR threshold are retained. Otherwise, "softly" developed inundated areas may convert to wetland or open water habitats. The model is run under both Protected Development (PD) and Allowed Migration (AM) scenarios to bound uncertainty in projected habitat outcomes with divergent future land management policies.

### 4.7 02_HabitatEvolutionModel_v202512.R --- Master Script

The master orchestration script serves as the primary entry point for full model execution. It reads the SLR scenario table, iterates through each projection year in sequence, and calls all sub-model scripts in the appropriate order: datum adjustment, marsh accretion, freshwater processing, ocean proximity, and habitat reclassification. For each projection year and SLR scenario, the script writes a classified habitat raster (`habitat_{year}.tif`) to the `outputs/` directory. These annual outputs constitute the primary deliverable of the Habitat Evolution Model and serve as inputs to the county-level change analysis reported in the HEM Change Summary (Table 2).

## 5. Model Outputs and Results Summary

### 5.1 Primary Output Products

For each year specified in the SLR scenario table, the Habitat Evolution Model produces a classified GeoTIFF raster containing predicted habitat type codes for the entire seven-county project area. Two parallel output sets are produced under the Protected Development (PD) and Allowed Migration (AM) land policy scenarios. Intermediate raster products, including accreted topography and freshwater masks, are also produced for quality review and diagnostic purposes by model practitioners.

Summary statistics of habitat area by county, HEM category, projection year, and SLR scenario are compiled and reported through an automated HTML reporting product accessible at <https://tbep-tech.github.io/tbcmp-analyze/hemchange.html>. This report is generated using R and rendered via GitHub Actions when updated analytical outputs become available.

### 5.2 Representative Model Results

Table 2 presents a representative selection of projected habitat areas (in hectares) for the Tampa Bay region under the modeled SLR scenarios, illustrating the magnitude and direction of anticipated habitat transitions. Results span the baseline condition (2025) through projections from 2050 to 2100 under the PD and AM land policy scenarios and the low and high accretion scenarios. As previously reported under accelerating sea level rise ([Sheehan et al. 2019](https://doi.org/10.1007/s13157-019-01137-y)), substantial changes in high marsh and seagrass habitats are expected, while mangrove and open water extents are projected to expand significantly in the near-to-mid term before declining at higher SLR increments.

**Table 2.** *Representative projected habitat areas (hectares) for the seven-county Tampa Bay TBCMP project area under modeled SLR scenarios.*

```{r}
#| label: tbl-results
#| tbl-cap: "Representative projected habitat areas (hectares) for the seven-county Tampa Bay TBCMP project area under modeled SLR scenarios."

results_df <- data.frame(
  `HEM Category` = c(
    "Subtidal - Tidal Flat",
    "Subtidal – Seagrass",
    "Mangroves",
    "High Marshes",
    "Juncus Marshes",
    "Salt Barrens",
    "Freshwater Wetlands"
  ),
  `Baseline (acres)` = c(
    "47,753",
    "460,171",
    "19,687",
    "47,908",
    "Not Defined",
    "1,205",
    "486,598"
  ),
  `2050 Projection (acres)` = c(
    "583 - 1,281",
    "174,844 - 182,980",
    "33,291 - 44,206",
    "30,791 - 41,188",
    "5,049 - 6,972",
    "15,491 - 17,899",
    "459,386 - 464,927"
  ),
  `2080 Projection (ha)` = c(
    "1,546 - 4,539",
    "134,317 - 181,532",
    "32,880 - 65,947",
    "7,705 - 40,384",
    "5,090 - 9,617",
    "13,106 - 20,351",
    "433,188 - 454,158"
  ),
  `2100 Projection (ha)` = c(
    "1,048 - 4,215",
    "117,658 - 189,042",
    "21,809 - 70,080",
    "3,966 - 45,059",
    "2,028 - 8,879",
    "9,771 - 19,595",
    "417,976 - 448,089"
  ),
  check.names = FALSE
)

knitr::kable(results_df, align = c("l", "c", "c", "c", "c"))
```

*Note: Values represent selected HEM categories aggregated across the project area. PD = Protected Development scenario; AM = Allowed Migration scenario. Full county-level results are available in the [online HEM Change Summary report](https://tbep-tech.github.io/tbcmp-analyze/hemchange.html).*

## 6. Technical Requirements and System Configuration

### 6.1 Software Dependencies

The Habitat Evolution Model requires R version 4.0 or higher and the following R packages:

- **terra** --- Primary raster processing library; replaces ArcGIS Spatial Analyst raster algebra functions.
- **sf** --- Spatial vector operations; replaces arcpy.analysis vector tools.
- **tidyverse** --- Data manipulation and workflow piping.
- **tidycensus** --- Programmatic access to US Census spatial boundary files.
- **curl** --- Programmatic data download from agency web endpoints.
- **here** --- Working directory management for reproducible file paths.
- **tbeptools** --- TBEP-specific data access and reporting utility functions.
- **dplyr** --- Data frame manipulation.

A minimum of 16 GB RAM is required to execute the model across the full seven-county project area, with 32 GB or more recommended for optimal performance. Out of memory errors within the R processing environment may be common on machines with less memory. GDAL must be installed on the host system, though this requirement is generally satisfied automatically by the `terra` package installation. Output rasters are written with LZW compression to manage file sizes for the large study area. Most of the rasters produced over the seven county area are too large to directly push to the GitHub repository and are therefore only stored locally (and subsequently ignored in .gitignore files). 

### 6.2 Translation from ArcGIS Model Builder Scripts to Open-Source R

A defining characteristic of this workflow is its complete translation from a proprietary ArcGIS ModelBuilder and ArcPython environment to the open-source R programming language. This translation was conducted to align the model with TBEP's open data science philosophy  and to enable version-controlled, automated, and reproducible execution through GitHub Actions. Table 3 summarizes the functional equivalences between ArcGIS operations and their R/terra counterparts implemented in this workflow.

**Table 3.** *Functional equivalences between ArcGIS/Python operations and R/terra implementations.*

```{r}
#| label: tbl-arcgis
#| tbl-cap: "Functional equivalences between ArcGIS/Python operations and R/terra implementations."

arcgis_df <- data.frame(
  `ArcGIS / Python Operation` = c(
    "Con()",
    "ExtractByMask()",
    "RasterToPolygon()",
    "PolygonToRaster()",
    "Buffer()",
    "MosaicToNewRaster()"
  ),
  `R / terra Equivalent` = c(
    "ifel()",
    "mask() or conditional ifel()",
    "as.polygons()",
    "rasterize()",
    "st_buffer() via sf package",
    "merge() or mosaic()"
  ),
  check.names = FALSE
)

knitr::kable(arcgis_df, align = c("l", "l"))
```

Key behavioral differences between the ArcGIS and R implementations include differences in NoData handling (ArcGIS uses -999 or system NoData; R/terra uses NA), raster algebra broadcasting (explicit vectorization via `ifel()` in R versus implicit broadcasting in ArcGIS), and vector operations (requiring the `sf` package in R). The workflow addresses these differences explicitly, with conditional checks for both -999 sentinel values and NA to ensure compatibility with legacy input data formats.

## 7. Quality Control and Automation

### 7.1 Codebase Quality Control

Quality control of the analytical codebase follows open-source software development guidelines as adopted by the TBEP ([rOpenSci 2024](https://doi.org/10.5281/zenodo.10797633)). Code is reviewed for redundancy, computational efficiency, and functional reproducibility. Automated software tests are scripted and executed weekly via GitHub Actions for each new analytical routine added to the repository. Tests verify that individual functions produce expected outputs under defined input conditions, ensuring that code modifications do not inadvertently alter model behavior. Identified defects and required corrections are documented using GitHub's issue tracking system, with each issue assigned to a responsible team member and linked to the specific file or code line requiring remediation.

### 7.2 Automation via GitHub Actions

Routine data extraction and analytical updates are automated through GitHub Actions workflows configured in the repository. When updated input datasets become available from source agencies, automated routines can be triggered to re-execute the data preparation and model runs, generating updated output rasters and refreshed HTML reporting products without manual intervention. This automation framework directly supports the annual update cycle described in the [TBCMP Technical Strategy](https://docs.google.com/document/d/1ol1cWYR-7QZDhsDrg3hRf3xOvu_7JMQs/edit?usp=sharing&ouid=107198491532889175628&rtpof=true&sd=true) and ensures that reporting products remain current with the most recent available land cover and environmental data.

### 7.3 Version Control and Documentation

All workflow scripts, data processing routines, and output products are maintained under Git version control, with the repository hosted publicly on the TBEP GitHub group page. Repository tags and commits provide a complete, time-stamped record of all workflow changes and dataset updates. Plain text README files accompany all data files and describe dataset provenance, processing steps, and metadata attributes following the Ecological Metadata Language (EML) standard. All data deliverables are structured in non-proprietary, standardized formats (.csv, .tif, .html) to ensure long-term accessibility and interoperability.

## References

[Beck, M.W., Raulerson, G.E., Burke, M.C., Whalen, J., Scolaro, S., & Sherwood, E.T. (2021). Data Management SOP for the Tampa Bay Estuary Program (Technical Report #12-21). Tampa Bay Estuary Program, St. Petersburg, Florida.](https://drive.google.com/file/d/1vO4B8DJATgCSV1qOxZz-kN6Uj1BrgNsg/view?usp=sharing)

[Beck, M.W., Robison, D.E., Raulerson, G.E., Burke, M.C., Saarinen, J., Sciarrino, C., Sherwood, E.T., & Tomasko, D.A. (2023). Addressing climate change and development pressures in an urban estuary through habitat restoration planning. *Frontiers in Ecology and Evolution*, 11, 1070266. <https://doi.org/10.3389/fevo.2023.1070266>](https://doi.org/10.3389/fevo.2023.1070266)

[Robison, D., Ries, T., Saarinen, J., Tomasko, D., & Sciarrino, C. (2020). Tampa Bay Estuary Program 2020 Habitat Master Plan Update (Technical Report #07-20). Prepared by Environmental Science Associates for the Tampa Bay Estuary Program. 136 pp. + appendix.](https://drive.google.com/file/d/1Hp0l_qtbxp1JxKJoGatdyuANSzQrpL0I/view?usp=drivesdk)

[rOpenSci, Anderson, B., Chamberlain, S., DeCicco, L., Gustavsen, J., Krystalli, A., Lepore, M., Mullen, L., Ram, K., Ross, N., Salmon, M., Vidoni, M., Riederer, E., Sparks, A., & Hollister, J. (2024). rOpenSci Packages: Development, Maintenance, and Peer Review (Version 0.9.0). Zenodo. <https://doi.org/10.5281/zenodo.10797633>](https://doi.org/10.5281/zenodo.10797633)

[Sheehan, K.L., Sherwood, E.T., Pribble, J.R., & Robbins, C.L. (2015). Development and Application of a Habitat Evolution Model to Assess Sea Level Rise Impacts in Tampa Bay, FL. *Wetlands*, 39, 305--320. <https://doi.org/10.1007/s13157-019-01137-y>](https://doi.org/10.1007/s13157-019-01137-y)

[Southwest Florida Water Management District. (2025). Land use land cover data. c1990--2023. Available at: <https://data-swfwmd.opendata.arcgis.com/>](https://data-swfwmd.opendata.arcgis.com)

[Tampa Bay Coastal Master Plan: Technical Strategy, Draft Version 1. (2025). Tampa Bay Regional Planning Council, Pinellas Park, Florida.](https://docs.google.com/document/d/1ol1cWYR-7QZDhsDrg3hRf3xOvu_7JMQs/edit?usp=sharing&ouid=107198491532889175628&rtpof=true&sd=true)
