---
title: "HEM Land Use Change Summary by Drainage Basin"
author: "Tampa Bay Estuary Program"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    code-tools: 
      source: https://github.com/tbep-tech/tbcmp-analyze/blob/main/docs/hemchange_basins.qmd
    css: tbcmp-brand.css
toc: true
toc-depth: 2
lightbox: true
execute:
  echo: false
  warning: false
---

```{r}
#| include: false
library(here)
library(tidyverse)
library(flextable)
library(ftExtra)
library(rlang)

# ---------------------------------------------------------------------------
# Inputs
#
# Basin summaries come from 03c_process_tiffs_basins.R, coastal summaries from
# 04b_HEM_Change_Summary_by_Basin.R Part B, and the 2025 baseline for both from
# 03d_process_baseline_basins.R.
# ---------------------------------------------------------------------------
basin_dir <- here("data/output/basin/")
coast_dir <- here("data/output/coastal/")

read_summaries <- function(dir, domain_label) {
  files <- list.files(dir, pattern = "_summary\\.csv$", full.names = TRUE)
  if (length(files) == 0) {
    return(tibble())
  }
  files |>
    lapply(read.csv) |>
    bind_rows() |>
    mutate(domain = domain_label)
}

# HEM recode. Labels are copied verbatim from hemchange.qmd so the county and
# basin documents stay directly comparable. They differ deliberately from
# ClassName in the raster attribute table ("High Marshes" vs "High Salt
# Marsh"); do not reconcile one without the other.
hem_recode <- function(value) {
  case_when(
    value %in% c(1100, 1200)                   ~ "Upland Developed - Hard",
    value %in% c(1800, 1820, 2100, 2200, 2400) ~ "Upland Developed - Soft",
    value %in% c(1900, 3200, 4100, 4400)       ~ "Upland Undeveloped",
    value %in% c(5200)                         ~ "Open Freshwater",
    value %in% c(5400)                         ~ "Open Water",
    value %in% c(6110, 6410)                   ~ "Freshwater Marsh & Wetlands",
    value %in% c(6120)                         ~ "Mangroves",
    value %in% c(6420)                         ~ "High Marshes",
    value %in% c(6425)                         ~ "Juncus Marshes",
    value %in% c(6600)                         ~ "Salt Barrens",
    value %in% c(7100)                         ~ "Beach - Dune",
    value %in% c(6510)                         ~ "Subtidal - Tidal Flats",
    value %in% c(9113)                         ~ "Subtidal - Seagrass",
    TRUE                                       ~ NA_character_
  )
}

combined <- bind_rows(
  read_summaries(basin_dir, "Drainage Basin"),
  read_summaries(coast_dir, "Coastal Waters")
) |>
  filter(value != 0 & !is.na(value)) |>
  mutate(
    hem_category = hem_recode(value),
    unit         = id,
    huc8         = case_when(
      domain == "Coastal Waters" ~ "Coastal",
      TRUE                       ~ as.character(basin_huc8)
    )
  ) |>
  filter(!is.na(hem_category))

# set colors and precision
cols      <- c('darkgreen', 'red')
precision <- 1

# ---------------------------------------------------------------------------
# Ensemble aggregation
#
# Anything summarised ABOVE the unit level has to be summed within each
# accretion x SLR member before mean() and sd() are taken. The county document
# does not need this because a county IS the unit; here a HUC8 or a domain is a
# rollup of many basins, and pooling their individual SDs would treat units as
# varying independently when in fact every basin in the rollup sees the same
# SLR curve. sd() over the vector of realised rollup totals is the spread of
# the number actually printed in the table.
#
# complete() fills zeros because a category absent from one member means zero
# acres, not a missing observation - without it mean() runs over a short,
# upward-biased vector for categories that blink in and out across scenarios.
# ---------------------------------------------------------------------------
ensemble_summary <- function(dat, group_cols) {

  member_cols <- c("land_policy", "accretion", "slr_scenario", "yr")

  dat |>
    summarize(
      sum_acres = sum(acres),
      .by = all_of(c(group_cols, member_cols, "hem_category"))
    ) |>
    complete(
      nesting(!!!syms(member_cols)),
      nesting(!!!syms(c(group_cols, "hem_category"))),
      fill = list(sum_acres = 0)
    ) |>
    summarize(
      mean_acres = mean(sum_acres),
      std_acres  = sd(sum_acres),
      .by = all_of(c(group_cols, "land_policy", "hem_category", "yr"))
    )
}

# ---------------------------------------------------------------------------
# Table construction
#
# Scenario columns are derived from the data rather than hard-coded, so a
# missing accretion or SLR run drops a column instead of erroring out of the
# pivot. Ordering is year ascending, PD before AM within each year, matching
# hemchange.qmd.
# ---------------------------------------------------------------------------
scen <- combined |>
  filter(land_policy != "baseline") |>
  distinct(land_policy, yr) |>
  arrange(yr, desc(land_policy)) |>
  mutate(key = paste(land_policy, yr, sep = "_"))

scen_keys <- scen$key
scen_yrs  <- unique(scen$yr)

build_tab <- function(dat, group_col) {

  gcol <- group_col

  totab <- dat |>
    ungroup() |>
    pivot_longer(cols = c(mean_acres, std_acres),
                 names_to = "statistic", values_to = "acres") |>
    unite("land_policy_yr", land_policy, yr, sep = "_")

  totabave <- totab |>
    filter(statistic == "mean_acres") |>
    select(-statistic) |>
    pivot_wider(names_from = land_policy_yr, values_from = acres,
                names_sort = TRUE, values_fill = 0)

  totabstd <- totab |>
    filter(statistic == "std_acres") |>
    select(-statistic) |>
    pivot_wider(names_from = land_policy_yr, values_from = acres,
                names_sort = TRUE, values_fill = 0)

  # Direction of change against baseline, one colour column per scenario
  totabcol <- totabave |>
    pivot_longer(cols = all_of(scen_keys),
                 names_to = "scenario", values_to = "mean_acres") |>
    mutate(
      col = sign(mean_acres - baseline_2025),
      col = case_when(col == 1 ~ cols[1], col == -1 ~ cols[2], TRUE ~ "black"),
      scenario = paste0(scenario, "_col")
    ) |>
    select(all_of(c(gcol, "hem_category", "scenario", "col"))) |>
    pivot_wider(names_from = scenario, values_from = col, names_sort = TRUE)

  totabave |>
    left_join(totabstd, by = c(gcol, "hem_category"),
              suffix = c("_mean", "_std")) |>
    pivot_longer(cols = c(ends_with("_mean"), ends_with("_std")),
                 names_to = c("land_policy", "yr", "statistic"),
                 names_pattern = "(.*)_(.*)_(.*)", values_to = "acres") |>
    mutate(
      acres = round(acres, precision),
      acres = case_when(yr == 2025 & statistic == "std" ~ NA_real_,
                        TRUE ~ acres)
    ) |>
    pivot_wider(names_from = statistic, values_from = acres) |>
    mutate(
      mean = case_when(!is.na(mean) ~ format(mean, big.mark = ",", scientific = FALSE),
                       TRUE ~ NA_character_),
      std  = case_when(!is.na(std)  ~ format(std,  big.mark = ",", scientific = FALSE),
                       TRUE ~ NA_character_)
    ) |>
    unite("mean_std_acres", mean, std, sep = " ± ", na.rm = TRUE) |>
    unite("policy_yr", land_policy, yr, sep = "_") |>
    mutate(policy_yr = factor(policy_yr, levels = c("baseline_2025", scen_keys))) |>
    pivot_wider(names_from = policy_yr, values_from = mean_std_acres,
                names_sort = TRUE, values_fill = "0") |>
    left_join(totabcol, by = c(gcol, "hem_category"), suffix = c("", "_col")) |>
    arrange(hem_category)
}

tab_fun <- function(totab, group_col, group_value) {

  sub_dat <- totab |>
    filter(.data[[group_col]] == group_value) |>
    select(-all_of(group_col))

  if (nrow(sub_dat) == 0) {
    return(flextable(data.frame(Note = "No data for this unit")) |> autofit())
  }

  col_keys <- c("hem_category", "baseline_2025", scen_keys)

  # Header row pairs the policies under each year. Built from scen_yrs so an
  # incomplete scenario set still produces a valid header.
  yr_header <- c("", "")
  for (y in scen_yrs) {
    yr_header <- c(yr_header, rep(as.character(y), sum(scen$yr == y)))
  }

  hdr_labels <- c("HEM Category", "Baseline",
                  sub("_\\d+$", "", scen_keys))
  names(hdr_labels) <- col_keys

  ft <- flextable(sub_dat, col_keys = col_keys) |>
    add_header_row(values = yr_header) |>
    set_header_labels(values = as.list(hdr_labels))

  # Merge each year's pair of policy columns in the top header row
  pos <- 3
  for (y in scen_yrs) {
    n <- sum(scen$yr == y)
    if (n > 1) ft <- merge_at(ft, i = 1, j = pos:(pos + n - 1), part = "header")
    pos <- pos + n
  }

  ft |>
    color(j = scen_keys,
          color = unlist(sub_dat[paste0(scen_keys, "_col")], use.names = FALSE)) |>
    autofit()
}

# ---------------------------------------------------------------------------
# Rollups
# ---------------------------------------------------------------------------
dat_domain  <- ensemble_summary(combined, "domain")
dat_total   <- ensemble_summary(combined |> mutate(all = "All"), "all")
dat_huc8    <- ensemble_summary(filter(combined, domain == "Drainage Basin"), "huc8")
dat_coastal <- ensemble_summary(filter(combined, domain == "Coastal Waters"), "unit")
dat_feat    <- ensemble_summary(filter(combined, domain == "Drainage Basin"), "basin_feat")
dat_unit    <- ensemble_summary(filter(combined, domain == "Drainage Basin"), "unit")

tab_domain  <- build_tab(dat_domain,  "domain")
tab_total   <- build_tab(dat_total,   "all")
tab_huc8    <- build_tab(dat_huc8,    "huc8")
tab_coastal <- build_tab(dat_coastal, "unit")
tab_feat    <- build_tab(dat_feat,    "basin_feat")
tab_unit    <- build_tab(dat_unit,    "unit")

# --- Tab labels -------------------------------------------------------------
# HUC8 tabs are labelled with the code plus the two largest basins inside them,
# derived from the data. Formal watershed names are not carried in the SWFWMD
# layer, and deriving the label means it cannot go stale or be wrong. To use
# formal names instead, populate huc8_names below as c("03100206" = "Tampa
# Bay", ...) and the label will switch over.
huc8_names <- c()

huc8_index <- combined |>
  filter(domain == "Drainage Basin", land_policy == "baseline") |>
  summarize(acres = sum(acres), .by = c(huc8, basin_name)) |>
  arrange(huc8, desc(acres)) |>
  summarize(
    top_basins = paste(head(basin_name, 2), collapse = ", "),
    acres      = sum(acres),
    n_basins   = n(),
    .by = huc8
  ) |>
  mutate(
    label = if (length(huc8_names)) {
      coalesce(unname(huc8_names[huc8]), huc8)
    } else {
      paste0(huc8, " \u2014 ", str_to_title(top_basins))
    }
  ) |>
  arrange(desc(acres))

coastal_index <- combined |>
  filter(domain == "Coastal Waters", land_policy == "baseline") |>
  summarize(acres = sum(acres), label = first(basin_name), .by = unit) |>
  arrange(desc(acres))

# Largest basins only - 465 basins carry data, far too many to tab through
n_top <- 12

basin_index <- combined |>
  filter(domain == "Drainage Basin", land_policy == "baseline") |>
  summarize(acres = sum(acres),
            label = paste0(str_to_title(first(basin_name)),
                           " (", first(basin_huc8), ")"),
            .by = unit) |>
  arrange(desc(acres)) |>
  head(n_top)

# --- Reconciliation ---------------------------------------------------------
recon <- combined |>
  filter(land_policy == "baseline") |>
  summarize(acres = sum(acres), .by = domain)

n_basins  <- n_distinct(filter(combined, domain == "Drainage Basin")$unit)
n_coastal <- n_distinct(filter(combined, domain == "Coastal Waters")$unit)

fmt <- function(x) format(round(x), big.mark = ",")
```

Habitat and Ecosystem Model (HEM) projections summarized by
**SWFWMD/FDEP drainage basin** rather than by county. Each table gives mean
acres across the accretion × sea level rise ensemble, ± one standard
deviation, under protect existing development (PD) and allowing habitats 
to migrate (AM) land policy. Values in <span style="color:darkgreen">**green**</span> 
are above the 2025 baseline; values in <span style="color:red">**red**</span> 
are below it.

A county-based version of these results is available in
[hemchange.qmd](hemchange.html).

::: {.callout-note collapse="true"}
## Coastal waters and why they are reported separately

The drainage basin layer covers `r fmt(recon$acres[recon$domain == "Drainage Basin"])`
acres of the project area. The remaining
`r fmt(recon$acres[recon$domain == "Coastal Waters"])` acres are open Gulf and
nearshore waters with no basin assignment, carried here as a separate
**Coastal Waters** domain derived by differencing the project extent against
the dissolved basin layer.

Reporting only the basins would silently drop that area, and it is
disproportionately Open Water and Subtidal - Seagrass — precisely where sea
level rise drives change. Note that Tampa Bay itself is *not* part of the
coastal residual; it is carried in the basin layer as a `BAY` feature, as are
Sarasota Bay and Boca Ciega Bay.

Totals: `r n_basins` drainage basins and `r n_coastal` coastal units,
`r fmt(sum(recon$acres))` acres combined.
:::

# Project area totals

::: {.panel-tabset .nav-pills}

## All areas

```{r}
tab_fun(tab_total, "all", "All")
```

## Drainage basins

```{r}
tab_fun(tab_domain, "domain", "Drainage Basin")
```

## Coastal waters

```{r}
tab_fun(tab_domain, "domain", "Coastal Waters")
```

:::

# By watershed (HUC8)

Drainage basins grouped by eight-digit hydrologic unit, ordered by area. Tab
labels name the two largest basins in each unit.

::: {.panel-tabset .nav-pills}

```{r}
#| output: asis
for (i in seq_len(nrow(huc8_index))) {
  cat("\n## ", huc8_index$label[i], "\n\n", sep = "")
  cat(knitr::knit_child(
    text = c(
      "```{r}",
      paste0("tab_fun(tab_huc8, 'huc8', '", huc8_index$huc8[i], "')"),
      "```"
    ),
    quiet = TRUE
  ), sep = "\n")
}
```

:::

# Coastal waters

Residual project area outside the drainage basin layer, attributed to the
county whose limits it falls within, with the remainder offshore.

::: {.panel-tabset .nav-pills}

```{r}
#| output: asis
for (i in seq_len(nrow(coastal_index))) {
  cat("\n## ", coastal_index$label[i], "\n\n", sep = "")
  cat(knitr::knit_child(
    text = c(
      "```{r}",
      paste0("tab_fun(tab_coastal, 'unit', '", coastal_index$unit[i], "')"),
      "```"
    ),
    quiet = TRUE
  ), sep = "\n")
}
```

:::

# By hydrologic feature type

Basins grouped by the `FEATURE` attribute of the source layer. `BAY` units
carry the open estuarine water that would otherwise be mistaken for part of
the coastal residual.

::: {.panel-tabset .nav-pills}

```{r}
#| output: asis
feat_index <- combined |>
  filter(domain == "Drainage Basin", land_policy == "baseline") |>
  summarize(acres = sum(acres), .by = basin_feat) |>
  arrange(desc(acres))

for (i in seq_len(nrow(feat_index))) {
  cat("\n## ", feat_index$basin_feat[i], "\n\n", sep = "")
  cat(knitr::knit_child(
    text = c(
      "```{r}",
      paste0("tab_fun(tab_feat, 'basin_feat', '", feat_index$basin_feat[i], "')"),
      "```"
    ),
    quiet = TRUE
  ), sep = "\n")
}
```

:::

# Largest individual basins

`r n_basins` basins carry data, too many to present individually. The
`r n_top` largest by baseline acreage are shown; per-basin values for the full
set are in `data/output/basin/`.

::: {.panel-tabset .nav-pills}

```{r}
#| output: asis
for (i in seq_len(nrow(basin_index))) {
  cat("\n## ", basin_index$label[i], "\n\n", sep = "")
  cat(knitr::knit_child(
    text = c(
      "```{r}",
      paste0("tab_fun(tab_unit, 'unit', '", basin_index$unit[i], "')"),
      "```"
    ),
    quiet = TRUE
  ), sep = "\n")
}
```

:::

# Notes on the numbers

**Standard deviations are computed at the level they are reported.** For any
rollup above a single basin, acres are summed within each accretion × sea
level rise combination first, and the standard deviation is taken across those
ensemble totals. Pooling the individual basins' standard deviations would
assume the basins vary independently, which they do not — every basin in a
watershed sees the same sea level curve — and would overstate the spread.

**Zero-filling.** Where a habitat category is absent from one ensemble member
it is treated as zero acres rather than as a missing observation, so means are
taken over the full set of members.

**Baseline** is the 2025 condition from `data/tbcmp_hem_filled.tif`, processed
through the same zonal routine as the scenarios.

**Boundary cells.** Basin summaries count whole cells assigned to the basin
containing the cell centre; the coastal summaries use partial-cell coverage
fractions. At 2 m resolution the difference is immaterial at these areas, but
it means basin and coastal acreages are not derived by identical methods.
